### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix typos. Clean-up solved \notes.
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

\[\stkc{
%% EnumU
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\EnumU}} 
\\
%% NilE
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\NilE}{\EnumU}} 
\qquad
%% ConsE
\Rule{\Gamma \vdash \Bhab{t}{\UId} \quad
      \Gamma \vdash \Bhab{e}{\EnumU}}
     {\Gamma \vdash \Bhab{\ConsE{t}{e}}{\EnumU}}
}\]

@@

Our first tentative is the following:

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{} ]})  \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} ???)}      \\
                                   \DIndx{\DUnit}                        \\
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]

So, we first pack an element $S$ of $\Set$. Then, we would like to
express a notion of recursive argument \emph{indexed} by $S$. Because
our presentation is entirely first-order so far, we are not able to
express this notion. We shall remedy to this situation.

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx$ that takes an indexing set $H$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx$ use the elements
of $H$ to index as many recursive arguments.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
\end{array}
}
\end{array}
}\]

@@

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
successfully describe the $\DSigma$ case, by a simple appeal to the
higher-order induction on $S$. The $\DHindx$ case consists in packing
the $H$ in $\Set$ with a recursive argument.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{S}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{x}{(\EnumT{e})} \switch{e}{(\LAM{\_} \Mu{\DescD})}{\pi^f}{x}
\]

@@

\[
\begin{array}{@{}ll}
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spi{e}{\LAM{\_} \Desc}}
                \PITEL{\V{x}}{\EnumT{e}} \To \Desc
\end{array}
\]

@@

\[
\Rule{\Gamma \Vdash \propag{\push{t}{\spiD{e}}}
                           {t'}}
     {\Gamma \Vdash \begin{array}{@{}l}
                        \propag{\push{t}{\EnumT{e} \To \Desc}}
                               {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
                    \end{array}
     }\;\mbox{t is $[]$ or $[a,b]$}
\]

@@

The catamorphism can be characterised as an induction on the
description $D$, with a non-dependent motive targeting $T$. Given a
node $xs$ and the induction hypotheses, the method ought to build an
element of $T$. Provided that we know how to make an element of
$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
us take a look at this jigsaw:

@@

\[\stk{
\cata : \PITEL{D}{\Desc}
           \PI{T}{\Set}
           (\descop{D}{T} \To T) \To 
           \Mu{D} \To T \\
\cata\: D\: T\: f \mapsto
  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:hs} f\: ???)
}\]

We are left with implementing the \(???\) program. Recall that we have
\(\Bhab{xs}{\descop{D}{\Mu{D}}}\) and
\(\Bhab{hs}{\All{D}{(\Mu{D})}{(\LAM{\_} T)}{xs}}\) at hand. Our goal
is to make an element of \(\descop{D}{T}\). Intuitively, $xs$ is of
the right shape, but its sub-elements are of the wrong type. On the
other hand, for each sub-element of $xs$, $hs$ gives us the
corresponding element in $T$. Hence the name ``induction
hypotheses''. Therefore, to construct an element of \(\descop{D}{T}\),
we replace the recursive components of \(xs\) by their counterpart in
\(hs\):

\[\stk{
\F{replace} : \stk{\PITEL{D}{\Desc}
                   \PITEL{X,Y}{\Set}\\
                   \PI{xs}{\descop{D}{X}} 
                   \All{D}{X}{(\LAM{\_}Y)}{xs} \To
                   \descop{D}{Y}} \\
\F{replace}\: \DUnit\:          X\: Y\: \Void\:          \Void          \mapsto \Void \\
\F{replace}\: (\DSigma{S}{D})\: X\: Y\: \pair{s}{xs}{}\: ys             \mapsto
    \pair{s}{\F{replace}\: {D~s}\: X\: Y\: xs\: ys}{}                                 \\
\F{replace}\: (\DIndx{D})\:     X\: Y\: \pair{x}{xs}{}\: \pair{y}{ys}{} \mapsto
    \pair{y}{\F{replace}\: D\: X\: Y\: xs\: ys}{}
}\]

Filling the \(???\) with \(\F{replace}\: D\: (\Mu{D})\: T\: xs\: hs\) closes the
problem. In the type theory, we have built a generic catamorphism. Any
data-type will now come equipped with this operation, for free.

@@

\[
\stk{
\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
                     \PI{\V{X}}{\Set} 
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
    \FMComposite     & F (\FMFreeMonad\: F\: X) \To \FMFreeMonad\: F\: X    
\end{array}
}
\]

@@

\[
\FMFreeMonadD\: F\: X\: Z \mapsto X + F Z
\]

@@

In our setting, the functor will be represented by a tagged
description. The free monad construction will take such tagged
description, a set $X$ of variables, and will compute the tagged
description of the corresponding free monad. Implementing this
function is surprisingly easy:

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{E}{D}{}}\:X \mapsto
    \pair{\ListCons{\DVar{}}{E}}
         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. Unfolding the
interpretation of this definition, we convince ourself that this
corresponds to the functor $\FMFreeMonadD$. The fix-point operation
ties the knot and gives us the full-blown free monad construction.

@@

\note{This is not quite Mu of a tagged description}

\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & \Mu{(\toDesc{\FreeMonad{D}{X}})} \To
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{Y}})}) \To
           \Mu{(\toDesc{\FreeMonad{D}{Y}})} 
\end{array} \\
\subst\: D\: X\: Y\: x\: \sigma \mapsto
  \cata\: (\toDesc{\FreeMonad{D}{X}})\: 
          (\Mu{(\toDesc{\FreeMonad{D}{Y}})})\: 
          ???\: 
          x
}\]

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{x}$ and replace them by $\sigma x$. Otherwise, it should simply
appeal to the induction hypothesis. This corresponds to the following
definition:

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{X}})}) \To
           \descop{\toDesc{\FreeMonad{D}{X}}}{\Mu{(\toDesc{\FreeMonad{D}{Y}})}} \To
           \Mu{(\toDesc{\FreeMonad{D}{Y}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives our
user the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the corresponding
monadic operation, \bind and \return. Whereas the \return is a trivial
variable former, the \bind operation corresponds to the substitution
of variables by terms. We have shown how to implement it, for any free
monad.

>>>>

\[\stkc{
%% EnumU
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\EnumU}} 
\\
%% NilE
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\NilE}{\EnumU}} 
\qquad
%% ConsE
\Rule{\Gamma \vdash \Bhab{\V{t}}{\UId} \quad
      \Gamma \vdash \Bhab{\V{e}}{\EnumU}}
     {\Gamma \vdash \Bhab{\ConsE{\V{t}}{\V{e}}}{\EnumU}}
}\]

@@

Our first tentative attempt is the following:

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{} ]})  \\
                           & \left[\begin{array}{l}
                                   \DUnit                                \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
                                   \DIndx{\DUnit}                        \\
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]

So, we first pack an element $\V{S}$ of $\Set$. Then, we would like to
express a notion of recursive argument \emph{indexed} by $\V{S}$. Because
our presentation is entirely first-order so far, we are not able to
express this notion. We shall remedy to this situation.

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx$ that takes an indexing set $\V{H}$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx$ use the elements
of $H$ to index as many recursive arguments.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx         & \PI{\V{H}}{\Set} \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
successfully describe the $\DSigma$ case, by a simple appeal to the
higher-order induction on $\V{S}$. The $\DHindx$ case consists in packing
the $\V{H}$ in $\Set$ with a recursive argument.

@@

\[\stk{
\DescD : \Desc \\
\begin{array}{@{}ll}
\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]}) \\
                           & \left[\begin{array}{l}
                                   \DUnit                                            \\
                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{e}} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{\V{x}}{(\EnumT{\V{e}})} \switch{\V{e}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
\]

@@

\[
\begin{array}{@{}ll}
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spi{\V{e}}{\LAM{\_} \Desc}}
                \PITEL{\V{x}}{\EnumT{\V{e}}} \To \Desc
\end{array}
\]

@@

\[
\Rule{\Gamma \Vdash \propag{\push{\V{t}}{\spiD{\V{e}}}}
                           {\V{t'}}}
     {\Gamma \Vdash \begin{array}{@{}l}
                        \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
                               {\\ \PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
                    \end{array}
     }\;\mbox{$\V{t}$ is $[]$ or $[\_,\_]$}
\]

@@

The catamorphism can be characterised as an induction on the
description $\V{D}$, with a non-dependent motive targeting $\V{T}$. Given a
node $\V{xs}$ and the induction hypotheses, the method ought to build an
element of $\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra $\V{f}$. Let
us take a look at this jigsaw:

@@

\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f\: \V{x}} \mapsto
  \F{induction}\: \V{D}\:\V{x}\: (\LAM{\_}\V{T})\: (\LAM{\V{xs}\:\V{hs}} \V{f}\: \SHED)
}\]

We are left with implementing filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\) at hand. Our goal
is to make an element of \(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of
the right shape, but its sub-elements are of the wrong type. On the
other hand, for each sub-element of $\V{xs}$, $\V{hs}$ gives us the
corresponding element in $\V{T}$. Hence the name ``induction
hypotheses''. Therefore, to construct an element of \(\descop{\V{D}}{\V{T}}\),
we replace the recursive components of \(\V{xs}\) by their counterpart in
\(\V{hs}\):

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto \Void \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}
}\]

Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\: (\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the
problem. In the type theory, we have built a generic catamorphism. Any
data-type will now come equipped with this operation, for free.

@@

\[
\stk{
\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
                     \PI{\V{X}}{\Set} 
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
\end{array}
}
\]

@@

\[
\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F} \V{Z}
\]

@@

In our setting, the functor will be represented by a tagged
description. The free monad construction will take such tagged
description, a set $\V{X}$ of variables, and will compute the tagged
description of the corresponding free monad. Implementing this
function is surprisingly easy:

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\ListCons{\DVar{}}{\V{E}}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep $\V{E}$ and $\V{D}$
as they were, hence leaving the functor unchanged. Unfolding the
interpretation of this definition, we convince ourself that this
corresponds to the functor $\FMFreeMonadD$. The fix-point operation
ties the knot and gives us the full-blown free monad construction.

@@

\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{x}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
          (\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})})\: 
          \SHED\: 
          \V{x}
}\]

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. Otherwise, it should simply
appeal to the induction hypothesis. This corresponds to the following
definition:

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
\end{array}
}\]

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives our
user the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the corresponding
monadic operation, \bind and \return. Whereas the \return is a trivial
variable former, the \bind operation corresponds to the substitution
of variables by terms. We have shown how to implement it, for any free
monad.
